\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.Rmd
add direct CV computation to speed comparisons in JSS draft paper

[^cv-by-programming]: \proglang{R} users with sophisticated programming skills would generally find it unchallenging to implement CV directly for specific applications. Even in this case, however, there's an argument for using a simple pre-programmed interface to CV, to minimize programming effort and to avoid mistakes.

@@

We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded of the CV packages mentioned in the preceding paragraph). Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
```{r rsample-example, cache=TRUE}
library(rsample)

@@

And here's how we can perform the same computation using the \code{cv()} function in the \pkg{cv} package:
```{r rsample-example-using-cv, cache=TRUE}
mod.attrition <- glm(mod_form, data=attrition, family=binomial)
(cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
          check.attributes=FALSE)
```

Let's compare the computational efficiency of the two implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
```{r compare-rsample-cv, cache=TRUE}
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury")$`CV crit`,
          )
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="hatvalues")$`CV crit`
          )

microbenchmark::microbenchmark(
  rsample={
    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
    rs_obj$accuracy <- map_dbl(rs_obj$results,
                               function(x) mean(x$correct))
  },
  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury"),
  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
                  method="hatvalues"),
  times=10
)
```
Computing time for \pkg{rsample} for this problem is similar to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.

>>>>

[^cv-by-programming]: \proglang{R} users with sophisticated programming skills would generally find it unchallenging to implement CV directly for specific applications (as we illustrate below). Even in this case, however, there's an argument for using a simple pre-programmed interface to CV, to minimize programming effort and to avoid mistakes.

@@

We illustrate with an example adapted from a vignette in the \pkg{rsample} package, which uses the \code{attrition} data set from the \pkg{modeldata} package [@Kuhn:2024] (and which is by a wide margin the most downloaded from the RStudio CRAN mirror of the CV packages mentioned in the preceding paragraph). Here, with minimal explanation, is how one would perform LOO CV for this example using \pkg{rsample}:[^k-fold-used-in-rsample]
```{r rsample-example, cache=TRUE}
library(rsample)

@@

In this case, it's simple (and, in our opinion, simpler) to program LOO CV directly. For example
```{r direct-loo-cv, cache=TRUE}
mod.attrition <- glm(mod_form, data=attrition, family=binomial)
n <- nrow(attrition)
yhat <- numeric(n)
for (i in 1:n){
  m <- update(mod.attrition, data=attrition[-i, ])
  yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
}
mean(((attrition$Attrition == "Yes") - round(yhat))^2)
```

And here's how we can perform the same computation using the \code{cv()} function in the \pkg{cv} package:
```{r rsample-example-using-cv, cache=TRUE}
(cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
          check.attributes=FALSE)
```

Let's compare the computational efficiency of the three implementations, also showing the use of \code{method="Woodbury"} and \code{method="hatvalues"} for \code{cv()}, which, in this example, produce numerically identical results to the default \code{method="exact"}:
```{r compare-rsample-cv, cache=TRUE}
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury")$`CV crit`,
          )
all.equal(cv.attrition$`CV crit`,
          cv(mod.attrition, k="loo", criterion=BayesRule,
             method="hatvalues")$`CV crit`
          )

microbenchmark::microbenchmark(
  rsample={
    rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
    rs_obj$accuracy <- map_dbl(rs_obj$results,
                               function(x) mean(x$correct))
  },
  direct={
    n <- nrow(attrition)
    yhat <- numeric(n)
    for (i in 1:n){
      m <- update(mod.attrition, data=attrition[-i, ])
      yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
    }
    mean(((attrition$Attrition == "Yes") - round(yhat))^2)
  },
  cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
  cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
             method="Woodbury"),
  cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
                  method="hatvalues"),
  times=10
)
```
Computing time for \pkg{rsample} for this problem is similar to direct computation of LOO CV and to \code{cv()} with \code{method="exact"}, but more than an order of magnitude slower than \code{cv()} with \code{method="Woodbury"}, and more than three orders of magnitude slower then \code{cv()} with \code{method="hatvalues"}.

### 2024-05-27 scratch/notes_john/JSS-article/JSS-article-3.tex
add direct CV computation to speed comparisons in JSS draft paper

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 968919
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.22048
bias-adjusted cross-validation criterion = 19.20805
95% CI for bias-adjusted CV criterion = (15.74739, 22.66871)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: microseconds
      expr        min         lq       mean     median         uq        max
 hatvalues    992.282   1149.681   1181.427   1199.188   1223.235   1356.485
  Woodbury  12297.663  12375.317  14678.447  12908.952  16017.675  22227.535
     naive 217204.347 222086.545 249652.690 250411.416 277112.440 279141.530
    cv.glm 379237.823 382052.063 401064.952 388823.726 435052.394 440563.573
 neval
    10
    10
    10
    10
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x142fa0f70>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

Our goal in implementing CV is to make it conveniently available to
\proglang{R} users with limited programming skills,\footnote{\proglang{R}
  users with sophisticated programming skills would generally find it
  unchallenging to implement CV directly for specific applications. Even
  in this case, however, there's an argument for using a simple
  pre-programmed interface to CV, to minimize programming effort and to
  avoid mistakes.} and hence to encourage its use. Towards this end, the
\pkg{cv} package provides a simple, consistent interface through the
\code{cv()} generic function, with specific methods for different
classes of standard \proglang{R} statistical models. Moreover, we have
tried to make writing methods for additional classes of statistical
models as simple and straightforward as possible. Because CV is
computationally intensive, we also aim for efficiency, by enabling
parallel computations generally, and by exploiting properties of
specific classes of statistical models (in particular, computations
based on hatvalues and the Woodbury matrix identity for linear and
generalized linear model). Finally, the package includes some unusual
features, such as general support for mixed-effects models and for
cross-validating complex model-specification procedures.

@@

We illustrate with an example adapted from a vignette in the
\pkg{rsample} package, which uses the \code{attrition} data set from the
\pkg{modeldata} package \citep{Kuhn:2024} (and which is by a wide margin
the most downloaded of the CV packages mentioned in the preceding
paragraph). Here, with minimal explanation, is how one would perform LOO
CV for this example using \pkg{rsample}:\footnote{The example in the
  \pkg{rsample} vignette uses 10-fold rather than LOO CV, but otherwise
  the code from the vignette is only trivially modified here.}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> mod.attrition <- glm(mod_form, data=attrition, family=binomial)
R> (cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: exact
criterion: BayesRule
cross-validation criterion = 0.1612245
bias-adjusted cross-validation criterion = 0.1612245
95% CI for bias-adjusted CV criterion = (0.1424194, 0.1800296)
full-sample criterion = 0.1612245 
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
+           check.attributes=FALSE)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

Let's compare the computational efficiency of the two implementations,
also showing the use of \code{method="Woodbury"} and
\code{method="hatvalues"} for \code{cv()}, which, in this example,
produce numerically identical results to the default
\code{method="exact"}:

\begin{CodeChunk}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury")$`CV crit`,
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="hatvalues")$`CV crit`
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   rsample={
+     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+     rs_obj$accuracy <- map_dbl(rs_obj$results,
+                                function(x) mean(x$correct))
+   },
+   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
+   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury"),
+   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                   method="hatvalues"),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Unit: milliseconds
         expr         min          lq        mean      median          uq
      rsample 5858.845429 5918.511745 6182.166132 6032.719008 6513.721496
     cv.exact 4823.821011 4879.622175 4986.028516 4973.975331 5051.786505
      cv.wood  103.815690  107.823973  112.389204  112.138341  117.148398
 cv.hatvalues    1.248655    1.299823    1.540665    1.587007    1.654432
         max neval
 6707.589832    10
 5305.047728    10
  121.483287    10
    1.916832    10
\end{CodeOutput}
\end{CodeChunk}

Computing time for \pkg{rsample} for this problem is similar to
\code{cv()} with \code{method="exact"}, but more than an order of
magnitude slower than \code{cv()} with \code{method="Woodbury"}, and
more than three orders of magnitude slower then \code{cv()} with
\code{method="hatvalues"}.

>>>>

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 716237
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.28027
bias-adjusted cross-validation criterion = 19.2646
95% CI for bias-adjusted CV criterion = (15.81196, 22.71725)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta # MSE, biased-corrected MSE
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: microseconds
      expr        min         lq       mean     median         uq        max
 hatvalues    984.369   1182.604   1188.963   1202.345   1220.242   1293.468
  Woodbury  12366.543  12508.690  13408.099  12769.347  12815.124  16566.050
     naive 216467.126 217748.581 225187.170 219534.685 222515.077 276207.898
    cv.glm 373527.917 380937.683 393430.301 382341.584 389266.341 443234.190
 neval
    10
    10
    10
    10
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> BayesRule
\end{CodeInput}
\begin{CodeOutput}
function(y, yhat) {
  if (!all(y %in% c(0, 1)))
    stop("response values not all 0 or 1")
  if (any(yhat < 0) ||
      any(yhat > 1))
    stop("fitted values outside of interval [0, 1]")
  yhat <- round(yhat)
  result <- mean(y != yhat) # proportion in error
  attr(result, "casewise loss") <- "y != round(yhat)"
  result
}
<bytecode: 0x126d8b808>
<environment: namespace:cv>
\end{CodeOutput}
\end{CodeChunk}

@@

Our goal in implementing CV is to make it conveniently available to
\proglang{R} users with limited programming skills,\footnote{\proglang{R}
  users with sophisticated programming skills would generally find it
  unchallenging to implement CV directly for specific applications (as
  we illustrate below). Even in this case, however, there's an argument
  for using a simple pre-programmed interface to CV, to minimize
  programming effort and to avoid mistakes.} and hence to encourage its
use. Towards this end, the \pkg{cv} package provides a simple,
consistent interface through the \code{cv()} generic function, with
specific methods for different classes of standard \proglang{R}
statistical models. Moreover, we have tried to make writing methods for
additional classes of statistical models as simple and straightforward
as possible. Because CV is computationally intensive, we also aim for
efficiency, by enabling parallel computations generally, and by
exploiting properties of specific classes of statistical models (in
particular, computations based on hatvalues and the Woodbury matrix
identity for linear and generalized linear model). Finally, the package
includes some unusual features, such as general support for
mixed-effects models and for cross-validating complex
model-specification procedures.

@@

We illustrate with an example adapted from a vignette in the
\pkg{rsample} package, which uses the \code{attrition} data set from the
\pkg{modeldata} package \citep{Kuhn:2024} (and which is by a wide margin
the most downloaded from the RStudio CRAN mirror of the CV packages
mentioned in the preceding paragraph). Here, with minimal explanation,
is how one would perform LOO CV for this example using
\pkg{rsample}:\footnote{The example in the \pkg{rsample} vignette uses
  10-fold rather than LOO CV, but otherwise the code from the vignette
  is only trivially modified here.}

@@

In this case, it's simple (and, in our opinion, simpler) to program LOO
CV directly. For example

\begin{CodeChunk}
\begin{CodeInput}
R> mod.attrition <- glm(mod_form, data=attrition, family=binomial)
R> n <- nrow(attrition)
R> yhat <- numeric(n)
R> for (i in 1:n){
+   m <- update(mod.attrition, data=attrition[-i, ])
+   yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
+ }
R> mean(((attrition$Attrition == "Yes") - round(yhat))^2)
\end{CodeInput}
\begin{CodeOutput}
[1] 0.1612245
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> (cv.attrition <- cv(mod.attrition, k="loo", criterion=BayesRule))
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: exact
criterion: BayesRule
cross-validation criterion = 0.1612245
bias-adjusted cross-validation criterion = 0.1612245
95% CI for bias-adjusted CV criterion = (0.1424194, 0.1800296)
full-sample criterion = 0.1612245 
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`, 1 - mean(rs_obj$accuracy), 
+           check.attributes=FALSE)
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\end{CodeChunk}

Let's compare the computational efficiency of the three implementations,
also showing the use of \code{method="Woodbury"} and
\code{method="hatvalues"} for \code{cv()}, which, in this example,
produce numerically identical results to the default
\code{method="exact"}:

\begin{CodeChunk}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury")$`CV crit`,
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> all.equal(cv.attrition$`CV crit`,
+           cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="hatvalues")$`CV crit`
+           )
\end{CodeInput}
\begin{CodeOutput}
[1] TRUE
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   rsample={
+     rs_obj$results <- map(rs_obj$splits, holdout_results, mod_form)
+     rs_obj$accuracy <- map_dbl(rs_obj$results,
+                                function(x) mean(x$correct))
+   },
+   direct={
+     n <- nrow(attrition)
+     yhat <- numeric(n)
+     for (i in 1:n){
+       m <- update(mod.attrition, data=attrition[-i, ])
+       yhat[i] <- predict(m, newdata=attrition[i, ], type="response")
+     }
+     mean(((attrition$Attrition == "Yes") - round(yhat))^2)
+   },
+   cv.exact=cv(mod.attrition, k="loo", criterion=BayesRule),
+   cv.wood=cv(mod.attrition, k="loo", criterion=BayesRule,
+              method="Woodbury"),
+   cv.hatvalues=cv(mod.attrition, k="loo", criterion=BayesRule,
+                   method="hatvalues"),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(rsample = {: less accurate nanosecond
times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
         expr         min          lq        mean      median         uq
      rsample 5707.408772 5711.011032 5810.464433 5828.739437 5853.64979
       direct 5187.215696 5206.152940 5281.503347 5230.005408 5348.24496
     cv.exact 4659.560948 4691.022749 4735.789317 4737.251540 4790.76759
      cv.wood   99.232628  104.541882  106.917914  106.051051  108.49436
 cv.hatvalues    1.187401    1.224875    2.751965    1.356813    1.56374
        max neval
 5999.09487    10
 5500.60457    10
 4811.15176    10
  119.72115    10
   15.35548    10
\end{CodeOutput}
\end{CodeChunk}

Computing time for \pkg{rsample} for this problem is similar to direct
computation of LOO CV and to \code{cv()} with \code{method="exact"}, but
more than an order of magnitude slower than \code{cv()} with
\code{method="Woodbury"}, and more than three orders of magnitude slower
then \code{cv()} with \code{method="hatvalues"}.
